---
title: "5CourseProject1"
author: "The Squire"
date: "Friday, July 17, 2015"
output: html_document
---

First I will load a package to use the unzip function and then download the file from the internet and create a folder in the working directory if it doesn't already exist.  Finally I will set my working directory to my project folder

```{r}
library(downloader)

if (!file.exists("./5CourseProject1")){
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download(fileURL, dest = "5CourseProject1.zip", mode = "wb")
        unzip("5CourseProject1.zip", exdir = "./5CourseProject1")
        dateDownloaded <- date()
}

```

Now I will calculate the sum of steps by date and display those totals

```{r}
dat <- read.csv("./5CourseProject1/activity.csv", na.strings = "NA")

dateSteps <- tapply(dat$steps, dat$date, sum)
dateSteps
```

Plot the frequency of days with steps in intervals of 5,000:

```{r}
hist(dateSteps, breaks = c(0, 5000, 10000, 15000, 20000, 25000), main = "Typical Daily Step Counts", xlab = "Total Steps", ylab = "Number of Days")
```

Display mean and median of the total number of steps taken per day

```{r}
mean(dateSteps, na.rm = TRUE)
median(dateSteps, na.rm = TRUE)
```

Calculate mean at each time interval, show time series plot, and display max average

```{r}
dat$interval <- as.factor(dat$interval)
intervalSteps <- tapply(dat$steps, dat$interval, mean, na.rm = TRUE)
plot(intervalSteps, type = 'l', main = "Average Steps per 5 Minute Interval", xlab = "Number of 5 Minute Intervals", ylab = "Average Steps Taken")
max(intervalSteps)
```

Calculate the number of NAs in steps measurements and then replace them with mean for that interval

```{r}
sum(is.na(dat$steps))

datClean <- dat
for (i in 1:length(datClean$steps)){
        if (is.na(datClean$steps[i])){
                a <- as.character(datClean$interval[i])
                datClean$steps[i] <- intervalSteps[a]
        }
}
```

Plot the frequency of days with steps in intervals of 5,000 and display mean and median for number of steps per day:

```{r}
dateCleanSteps <- tapply(datClean$steps, datClean$date, sum)
hist(dateCleanSteps, breaks = c(0, 5000, 10000, 15000, 20000, 25000), main = "Typical Daily Step Counts", xlab = "Total Steps", ylab = "Number of Days")

mean(dateCleanSteps)
median(dateCleanSteps)
```

Calculate day of the week and plot weekday versus weekend steps

```{r}
datTemp <- datClean
datTemp$date <- as.character(datTemp$date)
datTemp$date <- as.POSIXlt(datTemp$date)
for (i in 1:length(datTemp$date)){
        datTemp$day[i] = weekdays(datTemp$date[i])
        if ((datTemp$day[i] == "Saturday") | (datTemp$day[i] == "Sunday")){
                datTemp$daycat[i] = "weekend"
        } else {
                datTemp$daycat[i] = "weekday"
        }
}
datTemp$daycat = as.factor(datTemp$daycat)

library(ggplot2)

weekSteps <- tapply(datTemp$steps, datTemp$interval, mean)
qplot(interval, steps, data = datTemp, facets = daycat~., main = "Weekday versus Weekend Steps", xlab = "Number of 5 Minute Intervals", ylab = "Average Steps Taken") + geom_line()
```